A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-15 text/manuscript.Rmd
Update fig s4, split dan's plot into fig 7-9
Update fig ref in text

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregated into one or more
predominant classes. To our surprise, most cancer types (with some notable
exceptions) were not heavily biased in one class, and all well-represented
cancer types had tumors in every class (see **figure 3A**
and full tumor profiles in **supplementary figure S1**, see also interactive
media from **supplementary file 1 (html)**), suggesting that, in principle at
least, our novel pathway-disruption classification method identifies molecular
pathology largely independent of tissue-of-origin. As an
example of one type of cancer that does have a biased pathway profile,
pancreatic adenocarcinoma (PAAD) was predominantly found in class 8, which is
distinguished  by KRAS mutations among others (**supplementary figure S3**). 
But even PAAD comprises tumors from the nine remaining classes, meaning that 
patients suffering these tumors have potentially different underlying molecular
pathologies.

@@

Tumor staging is based on physico-pathological criteria, including tumor diameter,
which can vary greatly in importance between different tissues. Stage is used
clinically as a proxy for advancement toward a more deadly state and metastasis.
Given these assumptions, it is possible that more advanced tumors have common
pathway disruption profiles, and the UMAP which features a series of lobe-like
structures on a common backbone of tumor samples (**figure 2C**) could in
principle also reflect progression through a series of stages. In support of
this hypothesis, the backbone starts with a cluster of tumors (class 1) that has the fewest point mutations and culminates in a cluster (class 10) which
has nearly every pathway disrupted (**figure 4**). However, outside of class 10 we don't observe
an obvious trend in the overall mutation burden across the backbone of the UMAP (**figure
1H**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
classes represent advancement through stages, we projected staging data onto the
UMAP. Similar to tissue of origin and other categories of tumor, we did not
observe any bias among the stages to specific pathway disruption classes
(**figure 3D**), suggesting that stage is not an determining factor in pathway
disruption classes.

@@

To discover what pathways drive the UMAP clustering, we calculated enrichment
for each pathway _in-class_ relative to all other classes combined (see methods)
and ranked pathways for each class from highest to lowest enrichment. We
visualized the enrichment as a heatmap (**figure 5B**). Using this approach, we
identified 25 pathways highly enriched (enrichment score > 0.3, see methods) in
class 2, 9 pathways enriched in class 3, 24 pathways in class 4, 10 pathways in
class 5, 10 pathways in class 6, 24 pathways in class 7, 24 pathways in class 8
(for which most pathways are putatively disrupted), 10 pathways in class 9, and
0 pathways for clusters 1 and 10. Classes 7, 8, and 9 were highly concordant in
terms of pathway enrichment. To explore the specific pathways marking
each class, we projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4**). Classes 3 and 5 were distinguished by metabolic
pathways including RNA and protein biosynthesis (**supplementary figure S4**).
Similarly, class 4 was distinguished by mutations affecting regulation of DNA
and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
and "Nucleosome assembly"). Classes 7-9 have in common mutations in
extracellular, intracellular, and immune-related signaling pathways (see
**figure5C** and **supplementary figure S4**). Cluster 2 had the highest pathway
enrichment levels of the three, having mutations in hedgehog signaling,
"β-catenin degradation", "cellular response to hypoxia", "regulation of
cell cycle" and "apoptosis" among others.

@@

```{r, results='asis'}
cat("\\newpage")
```
#### Figure 6
```{r child = '../code/figure4.Rmd'}
#
```
#### Figure 7
```{r, results='asis'}
cat("\\newpage")
```

>>>>

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregated into one or more
predominant classes. To our surprise, most cancer types (with some notable
exceptions) were not heavily biased in one class, and all well-represented
cancer types had tumors in every class (see **figure 3A**
and full tumor profiles in **supplementary figure S1**, see also interactive
media from **supplementary file 1 (html)**), suggesting that, in principle at
least, our novel pathway-disruption classification method identifies molecular
pathology largely independent of tissue-of-origin. As an
example of one type of cancer that does have a biased pathway profile,
pancreatic adenocarcinoma (PAAD) was predominantly found in class 8, which is
distinguished  by KRAS mutations among others (**supplementary figure S5**). 
But even PAAD comprises tumors from the nine remaining classes, meaning that 
patients suffering these tumors have potentially different underlying molecular
pathologies.

@@

Tumor staging is based on physico-pathological criteria, including tumor diameter,
which can vary greatly in importance between different tissues. Stage is used
clinically as a proxy for advancement toward a more deadly state and metastasis.
Given these assumptions, it is possible that more advanced tumors have common
pathway disruption profiles, and the UMAP which features a series of lobe-like
structures on a common backbone of tumor samples (**figure 2C**) could in
principle also reflect progression through a series of stages. In support of
this hypothesis, the backbone starts with a cluster of tumors (class 1) that has the fewest point mutations and culminates in a cluster (class 10) which
has nearly every pathway disrupted (**figure 4**). However, outside of class 10 we don't observe
an obvious trend in the overall mutation burden across the backbone of the UMAP. 
Nonetheless, to test the hypothesis that the molecular-pathway disruption
classes represent advancement through stages, we projected staging data onto the
UMAP. Similar to tissue of origin and other categories of tumor, we did not
observe any bias among the stages to specific pathway disruption classes
(**figure 3D**), suggesting that stage is not an determining factor in pathway
disruption classes.

@@

To discover what pathways drive the UMAP clustering, we calculated enrichment
for each pathway _in-class_ relative to all other classes combined (see methods)
and ranked pathways for each class from highest to lowest enrichment. We
visualized the enrichment as a heatmap (**figure 5B**). Using this approach, we
identified 25 pathways highly enriched (enrichment score > 0.3, see methods) in
class 2, 9 pathways enriched in class 3, 24 pathways in class 4, 10 pathways in
class 5, 10 pathways in class 6, 24 pathways in class 7, 24 pathways in class 8
(for which most pathways are putatively disrupted), 10 pathways in class 9, and
0 pathways for clusters 1 and 10. Classes 7, 8, and 9 were highly concordant in
terms of pathway enrichment. To explore the specific pathways marking
each class, we projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4**). Classes 3 and 5 were distinguished by metabolic
pathways including RNA and protein biosynthesis (**supplementary figure S4**).
Similarly, class 4 was distinguished by mutations affecting regulation of DNA
and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
and "Nucleosome assembly"). Classes 7-9 have in common mutations in
extracellular, intracellular, and immune-related signaling pathways (see
**figure 5C** and **supplementary figure S4**). Cluster 2 had the highest pathway
enrichment levels of the three, having mutations in hedgehog signaling,
"β-catenin degradation", "cellular response to hypoxia", "regulation of
cell cycle" and "apoptosis" among others.

@@

```{r, results='asis'}
cat("\\newpage")
```
#### Figure 6
```{r child = '../code/figure4.Rmd'}
#
```

### 2020-08-15 text/survival.Rmd
Update fig ref in text

The degree to which a given cancer accelerates aging can be determined by the
$k_{tis}$ multiplier (**Figure 7F**) multiplied by the class specific modifier
(**Figure 7G**). Looking at the tissue specific multiplier, the least deadly was
prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
risk multiplier of between 15 and 22 relative to the background risk of death in
the population. This undoubtedly is influenced by the fact that relatively young
patients are affected by GBM and that it is extremely deadly. Among the
deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and
pancreatic (PAAD) cancers. Apart from these top 4 cancers which ranged from 6-22
in relative risk of age-adjusted death, the remaining cancer types ranged from
about 1 to 5 in magnitude. A rate of 1 represents patients who die relatively
close to the age predicted purely from their demographics independent of the
information that they have the given cancer.

Our estimates of tissue-specific class effects, in contrast, ranged from less
than 1 up to about 3 or 4, reflecting that some classes are either less deadly
or more deadly than other classes within each cancer types (**Figure 7G**). A
class specific rate of 1 represents typical rate for this tissue type. For
several cancers ( _e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell
lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely
indistinguishable from the prior, reflecting that either there were too few
mortalities in the data to make an estimate (as expected for PRAD and THCA) or
two few samples, period. We did not observe class-specific trends that held true
across cancer types, which could result from different cancers having different
standards of care for example. In support of this interpretation, we also tested
a factored model which considered both cancer type and each class independent of
cancer type (not shown). Though we were able to successfully fit this model, it
is a special case of the more general model where class and tissue can
independently affect longevity, and there is no reason to believe that class
specific effects would necessarily remain constant across tissue types given how
widely the patients vary across tissue types as well as the different drug and
surgical treatments typically used for each tissue.

>>>>

The degree to which a given cancer accelerates aging can be determined by the
$k_{tis}$ multiplier (**Figure 8**) multiplied by the class specific modifier
(**Figure 9**). Looking at the tissue specific multiplier, the least deadly was
prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
risk multiplier of between 15 and 22 relative to the background risk of death in
the population. This undoubtedly is influenced by the fact that relatively young
patients are affected by GBM and that it is extremely deadly. Among the
deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and
pancreatic (PAAD) cancers. Apart from these top 4 cancers which ranged from 6-22
in relative risk of age-adjusted death, the remaining cancer types ranged from
about 1 to 5 in magnitude. A rate of 1 represents patients who die relatively
close to the age predicted purely from their demographics independent of the
information that they have the given cancer.

Our estimates of tissue-specific class effects, in contrast, ranged from less
than 1 up to about 3 or 4, reflecting that some classes are either less deadly
or more deadly than other classes within each cancer types (**Figure 9**). A
class specific rate of 1 represents typical rate for this tissue type. For
several cancers ( _e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell
lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely
indistinguishable from the prior, reflecting that either there were too few
mortalities in the data to make an estimate (as expected for PRAD and THCA) or
two few samples, period. We did not observe class-specific trends that held true
across cancer types, which could result from different cancers having different
standards of care for example. In support of this interpretation, we also tested
a factored model which considered both cancer type and each class independent of
cancer type (not shown). Though we were able to successfully fit this model, it
is a special case of the more general model where class and tissue can
independently affect longevity, and there is no reason to believe that class
specific effects would necessarily remain constant across tissue types given how
widely the patients vary across tissue types as well as the different drug and
surgical treatments typically used for each tissue.
